Staged Abstract Interpreters

### 2019-09-08 paper/60_cases.tex
polish

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is a 0-CFA
for a small CPS language consisting of abstractions, applications, recursion, and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., @compCall@ and @compApp@).
Before refactoring to AC, the original functions take a store as a
regular argument. Now every function returns a value of type @CompAnalysis@,
which is a function value (i.e., closure) that takes and returns stores. After the
first run of the analysis, we have traversed the AST and obtained a single closure that
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now, we can invoke the compiled closure with an initial
store to complete the analysis.

>>>>

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is a 0-CFA
for a small CPS language consisting of abstractions, applications, recursion, and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled by different functions
(e.g., @compCall@ and @compApp@).
All original functions take a store as regular argument. After refactoring to
AC, every function returns a value of type @CompAnalysis@, which is a function
value that takes and returns stores. After the first run of the
analysis, we have traversed the AST and obtained a single closure that takes a
store and returns a store. Therefore, the ASTs are treated as static terms, and
the stores are dynamic terms. Finally, we may invoke the compiled closure with an
initial store to complete the analysis.

### 2019-09-08 paper/70_eval.tex
space
polish

To evaluate how performance can be improved by our staged abstract interpreters
approach, we scale our toy interpreter to a large subset of the Scheme language
and evaluate the analysis time against the unstaged versions. The result
shows that the staged versions are, on average, 10+ times as fast as the unstaged
version, as evidence of our \textit{abstraction without regret} approach.

\paragraph{Implementation and Evaluation Environment} We implement the abstract
interpreters for control-flow analysis, i.e., a value-flow analysis that
computes which lambda terms can be called at each call-site.
With other suitable abstract domains, the abstract interpreter
can also be extended to other analyses.
A front-end desugars Scheme to a small core language that makes the analyzer
easy to implement. We implement two common monovariant 0-CFA-like analyses: one
is equipped with store-widening (0CFA-SW for short), the other is not (0CFA
for short). It is traditionally expected that the store-widened analysis be faster than the
alternative.  Our prototype implementation currently generates Scala code; the
generated Scala code will be compiled and executed on the JVM, though it is possible to
generate C/C++ code in the future. In the experiments, we implement several
optimizations mentioned in Section \ref{staged_ds}, specifically, the selective
caching scheme and IR-level rewriting rules (for pairs, lists, sets, and maps)
that exploit partially-static data. These optimizations are useful to reduce
the size of generated code.

@@

\vspace{-1em}
\begin{figure}[h] \footnotesize
\begin{tabular}{@{}ll|lll|lll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ & unstaged   & staged    & $\frac{\text{unstaged}}{\text{staged}}$  \\
    \midrule
                        &      & \multicolumn{3}{c}{w/o store-widening}  &  \multicolumn{3}{c}{w/ store-widening}\\
    \midrule
    @fib@               & 32   & 3.288 ms   & 0.154 ms   & 21.33x      & 1.434 ms   & 0.098 ms  &  14.62x       \\
    @rsa@               & 451  & 238.171 ms & 23.333 ms  & 10.20x      & 11.977 ms  & 1.197 ms  &  10.00x       \\
    @church@            & 120  & 61.001 s   & 4.277 s    & 14.26x      & 2.338 ms   & 0.534 ms  &  4.37x        \\
    @fermat@            & 310  & 23.540 ms  & 2.885 ms   & 8.05x       & 7.146 ms   & 0.915 ms  &  7.81x        \\
    @mbrotZ@            & 331  & 665.456 ms & 66.070 ms  & 10.07x      & 11.008 ms  & 1.476 ms  &  7.45x        \\
    @lattice@           & 609  & 29.230 s   & 2.627 s    & 11.12x      & 16.432 ms  & 2.427 ms  &  6.76x        \\
    @kcfa-worst-16@     & 182  & 44.431 ms  & 3.211 ms   & 13.83x      & 4.425 ms   & 0.850 ms  &  5.20x        \\
    @kcfa-worst-32@     & 358  & 284.268 ms & 9.065 ms   & 31.35x      & 10.109 ms  & 1.661 ms  &  6.08x        \\
    @kcfa-worst-64@     & 710  & 2.165 s    & 0.0425 s   & 50.85x      & 23.269 ms  & 3.312 ms  &  7.02x        \\
    @solovay-strassen@  & 523  & 5.078 s    & 0.766 s    & 6.62x       & 18.757 ms  & 3.142 ms  &  5.96x       \\
    @regex@             & 550  & -          & -          & -           & 6.803 ms   & 1.088 ms  &  6.24x       \\
    @matrix@            & 1732 & -          & -          & -           & 85.611 ms  & 9.297 ms  &  9.20x       \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for monovariant control-flow analysis.} \label{evaluation_result}
\end{figure}

>>>>

To evaluate how performance can be improved by our staged abstract interpreters
approach, we scale our toy interpreter to support a large subset of the Scheme
language and evaluate the analysis time against the unstaged versions. The
result shows that the staged versions are, on average, 10+ times as fast as the
unstaged version, as evidence of our \textit{abstraction without regret}
approach.

\vspace{-5pt}
\paragraph{Implementation and Evaluation Environment} We implement the abstract
interpreters for control-flow analysis, i.e., a value-flow analysis that
computes which lambda terms can be called at each call-site.
With other suitable abstract domains, the abstract interpreter
can also be extended to other analyses.
A front-end desugars Scheme to a small core language that makes the analyzer
easy to implement. We implement two common monovariant 0-CFA-like analyses: one
is equipped with store-widening (0CFA-SW for short), the other is not (0CFA
for short). It is traditionally expected that the store-widened analysis be faster than the
alternative.  Our prototype implementation currently generates Scala code; the
generated Scala code will be compiled and executed on the JVM, though it is possible to
generate C/C++ code in the future. In the experiments, we implement several
optimizations mentioned in Section \ref{staged_ds}, specifically, the selective
caching scheme and IR-level rewriting rules (for pairs, lists, sets, and maps)
that exploit partially-static data. These optimizations are useful to reduce
the size of generated code.

@@

\begin{figure}[t] \footnotesize
\begin{tabular}{@{}ll|lll|lll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ & unstaged   & staged    & $\frac{\text{unstaged}}{\text{staged}}$  \\
    \midrule
                        &      & \multicolumn{3}{c}{w/o store-widening}  &  \multicolumn{3}{c}{w/ store-widening}\\
    \midrule
    @fib@               & 32   & 3.288 ms   & 0.154 ms   & 21.33x      & 1.434 ms   & 0.098 ms  &  14.62x       \\
    @rsa@               & 451  & 238.171 ms & 23.333 ms  & 10.20x      & 11.977 ms  & 1.197 ms  &  10.00x       \\
    @church@            & 120  & 61.001 s   & 4.277 s    & 14.26x      & 2.338 ms   & 0.534 ms  &  4.37x        \\
    @fermat@            & 310  & 23.540 ms  & 2.885 ms   & 8.05x       & 7.146 ms   & 0.915 ms  &  7.81x        \\
    @mbrotZ@            & 331  & 665.456 ms & 66.070 ms  & 10.07x      & 11.008 ms  & 1.476 ms  &  7.45x        \\
    @lattice@           & 609  & 29.230 s   & 2.627 s    & 11.12x      & 16.432 ms  & 2.427 ms  &  6.76x        \\
    @kcfa-worst-16@     & 182  & 44.431 ms  & 3.211 ms   & 13.83x      & 4.425 ms   & 0.850 ms  &  5.20x        \\
    @kcfa-worst-32@     & 358  & 284.268 ms & 9.065 ms   & 31.35x      & 10.109 ms  & 1.661 ms  &  6.08x        \\
    @kcfa-worst-64@     & 710  & 2.165 s    & 0.0425 s   & 50.85x      & 23.269 ms  & 3.312 ms  &  7.02x        \\
    @solovay-strassen@  & 523  & 5.078 s    & 0.766 s    & 6.62x       & 18.757 ms  & 3.142 ms  &  5.96x       \\
    @regex@             & 550  & -          & -          & -           & 6.803 ms   & 1.088 ms  &  6.24x       \\
    @matrix@            & 1732 & -          & -          & -           & 85.611 ms  & 9.297 ms  &  9.20x       \\
    \bottomrule
\end{tabular}
\vspace{-0.5em}
\caption{Evaluation result for monovariant control-flow analysis.} \label{evaluation_result}
\vspace{-2em}
\end{figure}
